\texttt{nuance}: Efficient detection of planets transiting active stars

### 2024-05-17 latex/paper.tex
fix: add proof reading corrections

\newcommand{\TODO}{\texttt{todo}}
\newcommand{\review}[1]{\textcolor{red}{#1}\textbf}
% \newcommand{\review}[1]{#1} 
\newcommand{\set}[1]{\{\,#1\,\}}
\newcommand{\footlink}[1]{\footnote{\url{#1}}}
\newcommand{\wtls}{\texttt{bi-weight+BLS}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\let\subsectionautorefname\sectionautorefname
\let\subsubsectionautorefname\sectionautorefname

@@

\begin{abstract}
    The detection of planetary transits in the light curves of active stars, featuring correlated noise in the form of stellar variability, remains a challenge. Depending on the noise characteristics, we show that the traditional technique that consists of detrending a light curve before searching for transits alters their signal-to-noise ratio, and hinders our capability to discover exoplanets transiting rapidly-rotating active stars. We present \nuancemethod{}, an algorithm to search for transits in light curves while simultaneously accounting for the presence of correlated noise, such as stellar variability and instrumental signals. We assess the performance of \nuancemethod{} on simulated light curves as well as on the TESS light curves of 438 rapidly-rotating M dwarfs. For each dataset, we compare our method to 5 commonly-used detrending techniques followed by a search with the Box-Least-Square algorithm. Overall, we demonstrate that \nuancemethod{} is the most performant method in 93\% of cases, leading to both the highest number of true positives and the lowest number of false positive detections. Although simultaneously searching for transits while modeling correlated noise is expected to be computationally expensive, we make our algorithm tractable and available as the \textsf{JAX}-powered Python package \href{https://github.com/lgrcia/nuance}{\nuancecode{}}, allowing its use on distributed environments and GPU devices. Finally, we explore the prospects offered by the \nuancemethod{} formalism, and its use to advance our knowledge of planetary systems around active stars, both using space-based surveys and sparse ground-based observations.
\end{abstract}

@@

Transiting exoplanets are keystone objects for the field of exoplanetary science, but detecting transits in light curves featuring stellar variability and instrumental signals remains a challenge (e.g. \citealt{Pont2006}, \citealt{Howell2016} or \citealt{Yaptangco2024}). For this reason, known transiting exoplanets tend to be found around quieter stars, or belong to the population of close-in giants whose transit signals dominate over stellar rotational variability \citep{Simpson2023}. However, transiting exoplanets around active stars are discoveries with significant scientific value. First, as younger stars are more active \citep{Skumanich1972}, being able to detect planets transiting active stars will favor the discovery of young planetary systems \citep[e.g.][]{Newton2022}. Second, as stellar variability may originate from surface active regions (such as starpots), transiting exoplanets can be used to map the photosphere of active stars \citep[e.g.][]{Morris2017}, benefiting both the study of stellar atmospheres and the concerning impact of their non-uniformity on planetary atmosphere retrievals \citep{rackham2018}. Overall, enabling the detection of transits in light curves with high levels of correlated noises will greatly benefit the study of terrestrial exoplanets around late M-dwarfs, usually observed at lower SNR and more likely to display photometric variability (e.g. \citealt{Murray2020} or \citealt{Petrucci2024}).

@@

Commonly used transit-search algorithms, such as the Box-Least-Square algorithm \citep[BLS,][]{bls} are capable of detecting transits in light curves containing only transit signals and white noise. Using this method, the simplest way to detect transits in a light curve featuring correlated noise (either astrophysical or instrumental), is to first clean it from nuisance signals before performing the search. This strategy is widely adopted by the community, both using physically-motivated systematic models like \cite{everest1, everest2}, or filtering techniques (\citealt{Jenkins2010}, \citealt{wotan}). However, when correlated noise starts resembling transits, this cleaning step, often called \textit{detrending}, is believed to degrade their detectability \cite[see subsection 4.3 of][]{wotan}. In this case, the only alternative to search for transits is to perform a full-fledged modeling of the light curve, including both transits and correlated noise, and to compute the likelihood of the data to the transit model on a wide parameter space, an approach largely avoided due to its intractable nature. Nonetheless, \cite{kovacs2016} ask: \textit{Periodic transit and variability search with simultaneous systematics filtering: Is it worth it?}. The latter study explores a handful of cases and generally discards the benefit of using a full-fledged approach.\\\\

@@

        \caption{The top plot shows the simulated flux time series of an active star drawn from the GP model described in \autoref{light_curves_simulations}. This dataset corresponds to an observation of 4 days with an exposure time of 2 minutes. The mean of this signal consists in a periodic transit signal of period $P=0.7$ days, duration $D=1.2$ hours and depth $\Delta=2\%$ (dark gray) plus an instrumental signal (blue). Correlated noise in the form of stellar variability is simulated with an SHO kernel of hyperparameters $\omega = \pi / 6 D$, $Q=45$ and $\sigma = \Delta$. Finally, white noise with a standard deviation of $0.1\%$ is added to the diagonal of the covariance matrix.}

@@

One way to quantify the detectability of a unique transit signal is to compute its signal-to-noise-ratio expressed as 
\begin{equation*}
  SNR= \frac{\Delta}{\sigma}\sqrt{n},
\end{equation*}
where $\Delta$ is the transit depth, $\sigma$ the measurements uncertainty (assuming homoscedasticity), and $n$ the number of points within transit. Although this metric is useful to assess the strength of the transit signal given a certain photometric precision, it does not account for the presence of correlated noise. However, instrumental and other astrophysical signals will necessarily affect the detectability of transits in realistic light curves \citep{Pont2006}. For this reason, the combined differential photometric precision (CDPP; \citealt{Jenkins2010}) metric was developed, and used in the context of the Kepler mission to assess the level of correlated noise in light curves, affecting the detectability of transits with a given duration. 
% Ultimately, this metric was essential to estimate the yield of the survey, and helped to achieve one of the primary goal of the mission: to measure the occurrence rate of Earth-size planets orbiting in the habitable zones of sun-like stars \citep{}.
The CDPP is computed by decomposing the data in the time-frequency domain using wavelets, and measures the significance of a distorted transit signal in the whitened data. For a given transit duration, the CDPP is then a measure of the noise remaining after filtering the light curve in each frequency band, taking into account the presence of non-stationary correlated noise of a given timescale (see \citealt{Jenkins2010} for more details). Hence, we can estimate the significance of the transit signal accounting for the presence of correlated noise as
\begin{equation}\label{eq:snr}
  SNR= \frac{\hat{\Delta}}{CDPP_D},
\end{equation}
where $CDPP_D$ is the CDPP computed for a given transit duration $D$ and $\bar{\Delta}$ is the transit depth after detrending. For simplicity, the CDPP is computed using the method from \cite{Gilliland2011} implemented in the \textsf{lightkurve} Python package \citep{Lightkurve} and $\hat{\Delta}$ is the depth obtained by solving \autoref{eq:linear_model} with the last column of $\bm{X}$ containing a normalized transit model.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{../workflows/plot_issues/figures/issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. An hour-long transit of depth 0.5\% is generated on top of white noise (standard deviation of 0.15\%) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a GP with an SHO kernel of hyperparameters $\omega=60$, $Q=0.5$ and $\sigma=\Delta/4$ . The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{light_curves_simulations}.}
        \label{fig:issue1}
    \end{centering}
\end{figure}
\autoref{fig:issue1} shows the SNR from \autoref{eq:snr} computed for a unique transit observed in the absence (grey) and presence of correlated noise (red). As illustrated in this figure, the presence of correlated noise strongly decreases the transit signal SNR, which would ultimately limit its detectability.

@@

In \autoref{fig:issue2}, we simulate a transit signal on top of which we add photometric stellar variability with different timescales, sampled from a GP with an SHO kernel described in \autoref{app_gp}. For each light curve, we reconstruct and detrend stellar variability in two ways: one using the widely-adopted Tukey's bi-weight filter, presented in \cite{tukey} and using the implementation from \textsf{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}; the other using the same GP from which the data has been sampled. We then estimate the resulting transit depth and compute the remaining transit SNR using \autoref{eq:snr}. \autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and intuitively suggests that this degradation due to detrending is strongly dependent on the correlated noise characteristics encountered.

@@

\autoref{fig:snr_detrend} shows that it exists an entire region of the $(\tau, \delta)$ parameter space for which the bi-weight detrending degrades transit SNR to the point of no detection ($SNR < 5$). Although being more robust, the same effect is observed when detrending with an optimal GP. Hence, detrending makes transit search blind to many systems, especially when the widely adopted bi-weight filter is used. We note that, outside this problematic parameter space, both techniques perform relatively well and such a degradation of the transit SNR should not be expected. Although this study should be extended to other detrending techniques, it highlights the need for a more informed transit search algorithm able to deal with correlated noise, at least if present in the form of stellar variability.

\newpage
\section{\textsf{nuance}}\label{nuance}
When searching for exoplanets using an indirect method, such as using transits or radial velocities, the initial detection is often as important as the follow-up observations, ultimately leading to the confirmation of the planetary candidate. For that reason, the final product of most transit search algorithms consists of a \textit{periodogram}: a 1D detection metric given a set of trial periods that allows to identify the presence of a potential candidate, and the period and epoch at which it should be followed up.\\\\ 
If we assume that a transit is defined by its period $P$, epoch $T_0$, duration $D$ and depth $\Delta$, we then wish to compute the likelihood of such a transit being present in the data for a set of periods $P$, leading to an interpretable periodogram. As we are interested in following up transits with specific parameters, such as a well-defined epoch, the likelihood we wish to compute must not be marginalized over all parameters other than $P$ but rather specifically computed at the maximum-likelihood parameters $\hat T_0$, $\hat D$ and $\hat \Delta$, leading to the \textit{profile likelihood}
\begin{equation}\label{eq:periodogram}
       \mathcal{Q}(P) = p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta),
\end{equation} 
where $\bm{f}$ is the data. In this section, we present \textit{nuance}, an algorithm to compute transit-search periodograms in a tractable way, and detect planetary transits in light curves containing correlated noise such as instrumental signals and photometric stellar variability. In \autoref{approach}, we explain how our approach requires two separate steps in order to remain tractable, and present these steps in \autoref{linear_search} and \autoref{periodic_search}, leading to the transit-search periodogram described in \autoref{periodogram}. Finally, we present the \nuancecode{} Python package in \autoref{package} and perform a control test of our implementation against BLS in \autoref{control}.

\subsection{The approach}\label{approach}
As in \autoref{light_curves_simulations}, let assume that $\bm{f}$ is a vector of size $N$ containing the flux of a star observed at times $\bm{t}$ such that
    \bm{f} \sim \mathcal{N}(\bm{X w}, \bm{\Sigma}),
i.e.\;that $\bm{f}$ is drawn from a GP of mean $\bm{Xw}$ and covariance $\bm{\Sigma}$. Again, we set the first $M-1$ columns of the $(N\times M)$ design matrix $\bm{X}$ as contemporaneous measurements and the last column as a normalized box-shaped transit of period $P$, epoch $T_0$ and duration $D$. This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given the presence of a periodic transit signal of period $P$, epoch $T_0$, duration $D$, and a mean linear model with coefficients $\bm{w}$ is \citep{Rasmussen2005}
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert P, T_0, D, \bm{w}) = -\frac{1}{2}(\bm{f}-\bm{Xw})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{Xw}) -  \frac{1}{2}\ln\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2\pi.
Given the linearity of the mean model at $P$, $T_0$ and $D$ fixed, this likelihood is maximized for the least-square parameters
\begin{equation}\label{eq:ls}
    \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \hspace{0.5cm} \text{with  uncertainties} \hspace{0.5cm} \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}.
% Using this expression, the depth of the transit corresponds to $\hat{\Delta} = \bm{\hat{w}}_M$ with the uncertainty $\hat{\sigma} = \bm{\hat{\sigma}}_{M,M}$, where M denotes the index of the last column of the design matrix $\bm{X}$, corresponding to the normal\ transit signal. 
As we are only interested in the transit depth, we will omit $\bm{w}$ in the reminder of this paper and simply write likelihood expressions depending on its last value $\Delta=\bm{w}_M$, such as $p(\bm{f} \vert P, T_
0, D, \Delta)$, where all other parameters of $\bm{w}$ are taken at their maximum-likelihood values. \\\\
Hence, given a period $P$, computing $\mathcal{Q}(P)$ boils down to a non-linear optimization involving several evaluations of the likelihood given in \autoref{eq:linear_search_ll}. While this is tractable for few periods, it becomes highly untractable for the large set of trial periods required to properly sample a transit-search periodogram (in the order of tens of thousands).\\\\
To remain tractable, \nuancemethod{} employs an extension of the strategy used by \cite{foreman2016}, which has significant intellectual overlap with the methods used by \cite{Aigrain2004} and \cite{Jenkins2010}. This approach separates the transit search into two components: the linear search and the periodic search. During the linear search, the likelihood $p(\bm{f} \vert T_0, D, \Delta)$ of a single non-periodic transit is computed for a grid of epochs and durations, each time solving for $\Delta$ linearly. Then, the periodic search consists of combining these likelihoods to compute the likelihood $p(\bm{f} \vert P, T_0, D, \Delta)$ of the data given a periodic transit signal for a range of periods $P$. These combined likelihoods yield $p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta)$, a transit-search periodogram on which the periodic transit detection is based. \nuancemethod{} differs from \cite{foreman2016} and other existing transit search algorithms as it models the covariance of the light curve with a GP, accounting for correlated noise (especially in the form of stellar variability) while keeping the model linear and tractable. This way, \nuancemethod{} searches for transits while, at the same time, modeling correlated noise, avoiding the separate detrending step that degrades transit signals SNR (see \autoref{issues}).\\\\
% We note that the approach employed by \nuancemethod{} (using the two steps proposed by \citealt{foreman2016}), shares similarities with the approach of \cite{Jenkins2010}, where a single event statistic is computed and combined into a multiple event statistics.

@@

t0s = time.copy()
# a range of 10 durations
Ds = np.linspace(0.01, 0.2, 10)
nu.linear_search(t0s, Ds)
\end{lstlisting}
Finally, the periodic search is run with
\begin{lstlisting}[language=Python]
# range of periods
periods = np.linspace(0.1, 5, 2000)
search = nu.periodic_search(periods)
\end{lstlisting}
From this \texttt{search} object, the best transiting candidate parameters can be computed (\textsf{search.best}), or the $\mathcal{Q}$ periodogram retrieved (\textsf{search.Q\_snr}), together with valuable information about the transit search. The \textsf{Nuance} object also provides methods to perform transit search on light curves from multi-planetary hosts, the advantage of \nuancemethod{} being that the linear search only needs to be performed once and reused for the search of several transiting candidates (cf.\;\autoref{toi540}). An extensive and maintained online documentation is provided at \href{https://nuance.readthedocs.io}{\texttt{nuance.readthedocs.io}}.

@@

To start testing \nuancecode{} against existing methods, a simple adimensional normalized light curve is simulated, consisting of pure white noise with a standard deviation of $5\times 10^{-4}$ spanning 6 days with an exposure time of 2 minutes.
From this signal, we produce 4000 light curves, each containing box-shaped transits with periods randomly sampled from 0.3 to 2.5 days, durations of 50 minutes, and depths randomly sampled to lead to transit SNRs ranging from 4 to 30.
% (i.e.\;depths from $2.6 \times 10^{-4}$ to $6.8 \times 10^{-4}$, using the simple model from \cite{protopapas} presented in \autoref{app_proto} with $c=500$). By design, these injected signals have an SNR ranging from 5 to 30 (with $\sigma_r = 0$ in \autoref{eq:snr}). 
For each light curve, transits are searched using two different tools: \textsf{nuance}, using its implementation from the Python package described in the previous section; and BLS, the Box-Least-Square algorithm from \cite{bls} (using \textsf{astropy}'s \textsf{BoxLeastSquares}\footlink{https://docs.astropy.org/en/stable/api/astropy.timeseries.BoxLeastSquares.html} implementation). \review{We note that the Transit-Least-Square algorithm \citep{tls}, that includes the effect of limb-darkening in the base transit template, could also be used here. However, as this effect have a negligible impact on the transit detection compared to the effect of correlated noise, we choose to make our comparisons with BLS only.}\\\\
For both methods, 3000 trial periods from 0.2 to 2.6 days are searched, with a single trial duration fixed to the unique known duration of 50 minutes. A transit signal is considered detected if the absolute difference between the injected and the recovered period is less than 0.01 day. To ease the detection criteria, orbital periods recovered at half or twice the injected ones (aka \textit{aliases}) are considered as being detected. For this reason, detected transit epochs are not considered (although manually vetted). Results from this \textit{injection-recovery} procedure are shown in \autoref{fig:control}.\\\\
These results demonstrate the qualitative match between the detection capabilities of \textsf{nuance} and \textsf{BLS} on light curves with no correlated noise, where the BLS algorithm should be optimal. Explaining the subtle differences observed between the two methods when only white noise is present is beyond the scope of this paper, and we will assume that any differences observed in the following sections are due to the different treatments of correlated noise.

@@

%  While this kernel is optimal, a comprehensive discussion on the use of non-optimal kernels is discussed in . 

>>>>

\newcommand{\TODO}{\texttt{todo}}
\newcommand{\review}[1]{#1}
% \newcommand{\review}[1]{#1} 
\newcommand{\set}[1]{\{\,#1\,\}}
\newcommand{\footlink}[1]{\footnote{\url{#1}}}
\newcommand{\wtls}{\texttt{bi-weight+BLS}}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
\let\subsectionautorefname\sectionautorefname
\let\subsubsectionautorefname\sectionautorefname

@@

\begin{abstract}
    The detection of planetary transits in the light curves of active stars, featuring correlated noise in the form of stellar variability, remains a challenge. Depending on the noise characteristics, we show that the traditional technique that consists of detrending a light curve before searching for transits alters their signal-to-noise ratio, and hinders our capability to discover exoplanets transiting rapidly-rotating active stars. We present \nuancemethod{}, an algorithm to search for transits in light curves while simultaneously accounting for the presence of correlated noise, such as stellar variability and instrumental signals. We assess the performance of \nuancemethod{} on simulated light curves as well as on the TESS light curves of 438 rapidly-rotating M dwarfs. For each dataset, we compare our method to 5 commonly-used detrending techniques followed by a search with the Box-Least-Squares algorithm. Overall, we demonstrate that \nuancemethod{} is the most performant method in 93\% of cases, leading to both the highest number of true positives and the lowest number of false positive detections. Although simultaneously searching for transits while modeling correlated noise is expected to be computationally expensive, we make our algorithm tractable and available as the \textsf{JAX}-powered Python package \href{https://github.com/lgrcia/nuance}{\nuancecode{}}, allowing its use on distributed environments and GPU devices. Finally, we explore the prospects offered by the \nuancemethod{} formalism, and its use to advance our knowledge of planetary systems around active stars, both using space-based surveys and sparse ground-based observations.
\end{abstract}

@@

Transiting exoplanets are keystone objects for the field of exoplanetary science, but detecting transits in light curves featuring stellar variability and instrumental signals remains a challenge (e.g. \citealt{Pont2006}, \citealt{Howell2016} or \citealt{Yaptangco2024}). For this reason, known transiting exoplanets tend to be found around quieter stars, or belong to the population of close-in giants whose transit signals dominate over stellar rotational variability \citep{Simpson2023}. However, transiting exoplanets around active stars are discoveries with significant scientific value. First, as younger stars are more active \citep{Skumanich1972}, being able to detect planets transiting active stars will favor the discovery of young planetary systems \citep[e.g.][]{Newton2022}. Second, as stellar variability may originate from surface active regions (such as starspots), transiting exoplanets can be used to map the photosphere of active stars \citep[e.g.][]{Morris2017}, benefiting both the study of stellar atmospheres and the concerning impact of their nonuniformity on planetary atmosphere retrievals \citep{rackham2018}. Overall, enabling the detection of transits in light curves with high levels of correlated noises will greatly benefit the study of terrestrial exoplanets around late M-dwarfs, usually observed at lower SNR and more likely to display photometric variability (e.g. \citealt{Murray2020} or \citealt{Petrucci2024}).

@@

Commonly used transit-search algorithms, such as the Box-Least-Squares algorithm \citep[BLS,][]{bls} are capable of detecting transits in light curves containing only transit signals and white noise. Using this method, the simplest way to detect transits in a light curve featuring correlated noise (either astrophysical or instrumental), is to first clean it from nuisance signals before performing the search. This strategy is widely adopted by the community, both using physically-motivated systematic models like \cite{everest1, everest2}, or filtering techniques (\citealt{Jenkins2010}, \citealt{wotan}). However, when correlated noise starts resembling transits, this cleaning step, often called \textit{detrending}, is believed to degrade their detectability \cite[see subsection 4.3 of][]{wotan}. In this case, the only alternative to search for transits is to perform a full-fledged modeling of the light curve, including both transits and correlated noise, and to compute the likelihood of the data to the transit model on a wide parameter space, an approach largely avoided due to its intractable nature. Nonetheless, \cite{kovacs2016} ask: \textit{Periodic transit and variability search with simultaneous systematics filtering: Is it worth it?}. The latter study explores a handful of cases and generally discards the benefit of using a full-fledged approach.\\\\

@@

        \caption{The top plot shows the simulated flux time series of an active star drawn from the GP model described in \autoref{light_curves_simulations}. This dataset corresponds to an observation of 4 days with an exposure time of 2 minutes. The mean of this signal consists of a periodic transit signal of period $P=0.7$ days, duration $D=1.2$ hours and depth $\Delta=2\%$ (dark gray) plus an instrumental signal (blue). Correlated noise in the form of stellar variability is simulated with an SHO kernel of hyperparameters $\omega = \pi / 6 D$, $Q=45$ and $\sigma = \Delta$. Finally, white noise with a standard deviation of $0.1\%$ is added to the diagonal of the covariance matrix.}

@@

One way to quantify the detectability of a unique transit signal is to compute its signal-to-noise-ratio expressed as 
\begin{equation*}
  SNR= \frac{\Delta}{\sigma}\sqrt{n},
\end{equation*}
where $\Delta$ is the transit depth, $\sigma$ the measurements uncertainty (assuming homoscedasticity), and $n$ the number of points within transit. Although this metric is useful to assess the strength of the transit signal given a certain photometric precision, it does not account for the presence of correlated noise. However, instrumental and other astrophysical signals will necessarily affect the detectability of transits in realistic light curves \citep{Pont2006}. For this reason, the combined differential photometric precision (CDPP; \citealt{Jenkins2010}) metric was developed, and used in the context of the Kepler mission to assess the level of correlated noise in light curves, affecting the detectability of transits with a given duration. 
% Ultimately, this metric was essential to estimate the yield of the survey, and helped to achieve one of the primary goal of the mission: to measure the occurrence rate of Earth-size planets orbiting in the habitable zones of sun-like stars \citep{}.
The CDPP is computed by decomposing the data in the time-frequency domain using wavelets, and measures the significance of a distorted transit signal in the whitened data. For a given transit duration, the CDPP is then a measure of the noise remaining after filtering the light curve in each frequency band, taking into account the presence of nonstationary correlated noise of a given timescale (see \citealt{Jenkins2010} for more details). Hence, we can estimate the significance of the transit signal accounting for the presence of correlated noise as
\begin{equation}\label{eq:snr}
  SNR= \frac{\hat{\Delta}}{CDPP_D},
\end{equation}
where $CDPP_D$ is the CDPP computed for a given transit duration $D$ and $\hat{\Delta}$ is the transit depth after detrending. For simplicity, the CDPP is computed using the method from \cite{Gilliland2011} implemented in the \textsf{lightkurve} Python package \citep{Lightkurve} and $\hat{\Delta}$ is the depth obtained by solving \autoref{eq:linear_model} with the last column of $\bm{X}$ containing a normalized transit model.
\begin{figure}[H]
    \begin{centering}
        \includegraphics[width=0.6\linewidth]{../workflows/plot_issues/figures/issue1.pdf}
        \caption{Illustration of the effect of correlated noise on a single transit SNR. An hour-long transit of depth 0.5\% is generated on top of white noise (standard deviation of 0.15\%) as part of a 24-hours observation with an exposure time of 1 minute (top). Then, in the bottom plot, correlated noise is added, generated using a GP with an SHO kernel of hyperparameters $\omega=60$, $Q=0.5$ and $\sigma=\Delta/4$ . The SNR on the right of each light curve is computed using \autoref{eq:snr}. Models used to simulate these data are provided in \autoref{light_curves_simulations}.}
        \label{fig:issue1}
    \end{centering}
\end{figure}
\autoref{fig:issue1} shows the SNR from \autoref{eq:snr} computed for a unique transit observed in the absence (gray) and presence of correlated noise (red). As illustrated in this figure, the presence of correlated noise strongly decreases the transit signal SNR, which would ultimately limit its detectability.

@@

In \autoref{fig:issue2}, we simulate a transit signal on top of which we add photometric stellar variability with different timescales, sampled from a GP with an SHO kernel described in \autoref{app_gp}. For each light curve, we reconstruct and detrend stellar variability in two ways: one using the widely-adopted Tukey's bi-weight filter, presented in \cite{tukey} and using the implementation from \textsf{wõtan}\footnote{\href{https://github.com/hippke/wotan}{https://github.com/hippke/wotan}} \citep{wotan}; the other using the same GP from which the data have been sampled. We then estimate the resulting transit depth and compute the remaining transit SNR using \autoref{eq:snr}. \autoref{fig:issue2} clearly shows the effect of both detrending techniques on transits SNR, and intuitively suggests that this degradation due to detrending is strongly dependent on the correlated noise characteristics encountered.

@@

\autoref{fig:snr_detrend} shows that there exists an entire region in the $(\tau, \delta)$ parameter space for which the bi-weight detrending degrades transit SNR to the point of no detection ($SNR < 5$). Although being more robust, the same effect is observed when detrending with an optimal GP. Hence, detrending makes transit search blind to many systems, especially when the widely adopted bi-weight filter is used. We note that, outside this problematic parameter space, both techniques perform relatively well and such a degradation of the transit SNR should not be expected. Although this study should be extended to other detrending techniques, it highlights the need for a more informed transit search algorithm able to deal with correlated noise, at least if present in the form of stellar variability.

\newpage
\section{\textsf{nuance}}\label{nuance}
When searching for exoplanets using an indirect method, such as using transits or radial velocities, the initial detection is often as important as the follow-up observations, ultimately leading to the confirmation of the planetary candidate. For that reason, the final product of most transit search algorithms consists of a \textit{periodogram}: a 1D detection metric given a set of trial periods that allows to identify the presence of a potential candidate, and the period and epoch at which it should be followed up.\\\\ 
If we assume that a transit is defined by its period $P$, epoch $T_0$, duration $D$ and depth $\Delta$, we then wish to compute the likelihood of such a transit being present in the data for a set of periods $P$, leading to an interpretable periodogram. As we are interested in following up transits with specific parameters, such as a well-defined epoch, the likelihood we wish to compute must not be marginalized over all parameters other than $P$ but rather specifically computed at the maximum-likelihood parameters $\hat T_0$, $\hat D$ and $\hat \Delta$, leading to the \textit{profile likelihood}
\begin{equation}\label{eq:periodogram}
       \mathcal{Q}(P) = p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta),
\end{equation} 
where $\bm{f}$ denotes the data. In this section, we present \textit{nuance}, an algorithm to compute transit-search periodograms in a tractable way, and detect planetary transits in light curves containing correlated noise such as instrumental signals and photometric stellar variability. In \autoref{approach}, we explain how our approach requires two separate steps in order to remain tractable, and present these steps in \autoref{linear_search} and \autoref{periodic_search}, leading to the transit-search periodogram described in \autoref{periodogram}. Finally, we present the \nuancecode{} Python package in \autoref{package} and perform a control test of our implementation against BLS in \autoref{control}.

\subsection{The approach}\label{approach}
As in \autoref{light_curves_simulations}, let assume that $\bm{f}$ is a vector of size $N$ containing the flux of a star observed at times $\bm{t}$ such that
\begin{equation*}
    \bm{f} \sim \mathcal{N}(\bm{X w}, \bm{\Sigma}),
\end{equation*}
i.e.\;that $\bm{f}$ is drawn from a GP of mean $\bm{Xw}$ and covariance $\bm{\Sigma}$. Again, we set the first $M-1$ columns of the $(N\times M)$ design matrix $\bm{X}$ as contemporaneous measurements and the last column as a normalized box-shaped transit of period $P$, epoch $T_0$ and duration $D$. This way, the transit signal is part of the linear model and its depth $\Delta$ can be solved linearly. Under this assumption, the log-likelihood of the data given the presence of a periodic transit signal of period $P$, epoch $T_0$, duration $D$, and a mean linear model with coefficients $\bm{w}$ is \citep{Rasmussen2005}
\begin{equation} \label{eq:linear_search_ll}
    \ln p(\bm{f} \vert P, T_0, D, \bm{w}) = -\frac{1}{2}(\bm{f}-\bm{Xw})^T\bm{\Sigma}^{-1}(\bm{f}-\bm{Xw}) -  \frac{1}{2}\ln\vert\bm{\Sigma}\vert - \frac{N}{2}\ln 2\pi.
\end{equation}
Given the linearity of the mean model at $P$, $T_0$ and $D$ fixed, this likelihood is maximized for the least-square parameters
\begin{equation}\label{eq:ls}
    \bm{w} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}\bm{X}^T\bm{\Sigma}^{-1}\bm{f} \hspace{0.5cm} \text{with  uncertainties} \hspace{0.5cm} \bm{\sigma} = (\bm{X}^T\bm{\Sigma}^{-1}\bm{X})^{-1}.
\end{equation}
As we are only interested in the transit depth, we will omit $\bm{w}$ in the remainder of this paper and simply write likelihood expressions depending on its last value $\Delta=\bm{w}_M$, such as $p(\bm{f} \vert P, T_
0, D, \Delta)$, where all other parameters of $\bm{w}$ are taken at their maximum-likelihood values. \\\\
Hence, given a period $P$, computing $\mathcal{Q}(P)$ boils down to a nonlinear optimization involving several evaluations of the likelihood given in \autoref{eq:linear_search_ll}. While this is tractable for few periods, it becomes highly untractable for the large set of trial periods required to properly sample a transit-search periodogram (in the order of tens of thousands).\\\\
To remain tractable, \nuancemethod{} employs an extension of the strategy used by \cite{foreman2016}, which has significant intellectual overlap with the methods used by \cite{Aigrain2004} and \cite{Jenkins2010}. This approach separates the transit search into two components: the linear search and the periodic search. During the linear search, the likelihood $p(\bm{f} \vert T_0, D, \Delta)$ of a single nonperiodic transit is computed for a grid of epochs and durations, each time solving for $\Delta$ linearly. Then, the periodic search consists of combining these likelihoods to compute the likelihood $p(\bm{f} \vert P, T_0, D, \Delta)$ of the data given a periodic transit signal for a range of periods $P$. These combined likelihoods yield $p(\bm{f} \vert P, \hat T_0 ,\hat D, \hat \Delta)$, a transit-search periodogram on which the periodic transit detection is based. \nuancemethod{} differs from \cite{foreman2016} and other existing transit search algorithms as it models the covariance of the light curve with a GP, accounting for correlated noise (especially in the form of stellar variability) while keeping the model linear and tractable. This way, \nuancemethod{} searches for transits while, at the same time, modeling correlated noise, avoiding the separate detrending step that degrades transit signals SNR (see \autoref{issues}).\\\\
% We note that the approach employed by \nuancemethod{} (using the two steps proposed by \citealt{foreman2016}), shares similarities with the approach of \cite{Jenkins2010}, where a single event statistic is computed and combined into a multiple event statistics.

@@

t0s = time.copy()
# a range of 10 durations
Ds = np.linspace(0.01, 0.2, 10)
nu.linear_search(t0s, Ds)
\end{lstlisting}
Finally, the periodic search is run with
\begin{lstlisting}[language=Python]
# range of periods
periods = np.linspace(0.1, 5, 2000)
search = nu.periodic_search(periods)
\end{lstlisting}
From this \texttt{search} object, the best transiting candidate parameters can be computed (\textsf{search.best}), or the $\mathcal{Q}$ periodogram retrieved (\textsf{search.Q\_snr}), together with valuable information about the transit search. The \textsf{Nuance} object also provides methods to perform transit search on light curves from multi-planetary hosts, the advantage of \nuancemethod{} being that the linear search only needs to be performed once and reused for the search of several transiting candidates (see \autoref{toi540}). An extensive and maintained online documentation is provided at \href{https://nuance.readthedocs.io}{\texttt{nuance.readthedocs.io}}.

@@

To start testing \nuancecode{} against existing methods, a simple adimensional normalized light curve is simulated, consisting of pure white noise with a standard deviation of $5\times 10^{-4}$ spanning 6 days with an exposure time of 2 minutes.
From this signal, we produce 4000 light curves, each containing box-shaped transits with periods randomly sampled from 0.3 to 2.5 days, durations of 50 minutes, and depths randomly sampled to lead to transit SNRs ranging from 4 to 30.
% (i.e.\;depths from $2.6 \times 10^{-4}$ to $6.8 \times 10^{-4}$, using the simple model from \cite{protopapas} presented in \autoref{app_proto} with $c=500$). By design, these injected signals have an SNR ranging from 5 to 30 (with $\sigma_r = 0$ in \autoref{eq:snr}). 
For each light curve, transits are searched using two different tools: \textsf{nuance}, using its implementation from the Python package described in the previous section; and BLS, the Box-Least-Squares algorithm from \cite{bls} (using \textsf{astropy}'s \textsf{BoxLeastSquares}\footlink{https://docs.astropy.org/en/stable/api/astropy.timeseries.BoxLeastSquares.html} implementation). \review{We note that the Transit-Least-Squares algorithm \citep{tls}, that includes the effect of limb-darkening in the base transit template, could also be used here. However, as this effect has a negligible impact on transit detection compared to the effect of correlated noise, we choose to make our comparisons with BLS only.}\\\\
For both methods, 3000 trial periods from 0.2 to 2.6 days are searched, with a single trial duration fixed to the unique known duration of 50 minutes. A transit signal is considered detected if the absolute difference between the injected and the recovered period is less than 0.01 day. To ease the detection criteria, orbital periods recovered at half or twice the injected ones (aka \textit{aliases}) are considered as being detected. For this reason, detected transit epochs are not considered (although manually vetted). Results from this \textit{injection-recovery} procedure are shown in \autoref{fig:control}.\\\\
These results demonstrate the qualitative match between the detection capabilities of \textsf{nuance} and \textsf{BLS} on light curves with no correlated noise, where the BLS algorithm should be optimal. Explaining the subtle differences observed between the two methods when only white noise is present is beyond the scope of this paper, and we will assume that any differences observed in the following sections are due to the different treatments of correlated noise.

@@
